Thesis: draft

### 2021-05-27 draft.tex
latest update

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss the results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, in order to investigate the interplay between filtering algorithms and network structure.

@@

\begin{figure}[H]
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_stubbornness_SBM_frac_compl_stubb_PR_10x8.png}
	\captionsetup{format=plain}
	\caption[Formation of echo chambers versus fraction of stubborn actors for the PR filtering algorithm and balanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The black dotted line represents a constant amount of echo chambers over time. The PR filtering algorithm is used. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}
\label{echo_vs_frac_complStubb}
\end{figure}
\noindent
The ER model shows no appearance of echo chambers. Introducing a fraction of stubborn actors does not change this. The SBM with low modularity behaves similar as the ER model. The data points for the SBM with low modularity lie behind the ER data points and are not visible in Fig. \ref{echo_vs_frac_complStubb}. The other networks (WS, SBM-WS and SBM with high modularity) all show qualitatively the same behavior. If an increasing fraction of stubborn actors is introduced, the formation of echo chambers decreases.\\
\newline
The WS and SBM-WS networks have high clustering. For a small fraction of stubborn actors they show the largest echo chambers. The SBM with high modularity has smaller echo chambers than networks with high clustering, but still has some formation of echo chambers for $f_s \leqslant 0.1$. These results confirm that community structure enhances the formation of echo chambers compared to random networks, however not as significant as networks with high clustering.

@@

  \caption[Formation of echo chambers versus fraction of stubborn actors for the PR and REC filtering algorithms and unbalanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for unbalanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. The $y$-axis shows the total number of like-minded neighborhoods (sum of the number of echo chambers around opinion A and the number of echo chambers around opinion B) at $t=500$ divided by the total number of like-minded neighborhoods at $t=0$. The black dotted line represents a constant amount of echo chambers over time. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$; these are small and not visible.}

@@

  \caption[Formation of echo chambers versus the resistance parameter of the actors, $r$, for unbalanced starting conditions. The PR and REC filtering algorithms are used.]{Formation of echo chambers versus the resistance parameter of the actors, $r$, for unbalanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). The black dotted line represent a constant amount of echo chambers over time. The $y$-axis shows the total number of like-minded neighborhoods (sum of the number of echo chambers around opinion A and the number of echo chambers around opinion B) at $t=500$ divided by the total number of like-minded neighborhoods at $t=0$. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$; these are small and not visible.}

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. $10 \%$ and $50 \%$ of the communities is given a single opinion, opinion A, while the remaining communities get an initial $50/50$ opinion distribution. The community distributed case is compared to the corresponding randomly distributed case ($55/45$ or $75/25$) for both a high modularity and a low modularity configuration. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. First row: $f_\text{A} = 0.1$: \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Second row: $f_\text{A} = 0.5$: \textbf{(c)} REC filtering algorithm; \textbf{(d)} PR filtering algorithm. The community distributed case is compared to the randomly distributed case for both a high and a low modularity configuration. The black dotted line represents a constant amount of echo chambers over time. The distribution is normalized by dividing for the same quantity computed at $t=0$. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}

@@

The formation of echo chambers versus the fraction of stubborn actors, $f_s$, in the Last.fm on-line social network is studied. Results are compared to results from network models, all with $N = 8003$: (i) WS model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM with 53 communities of size 151 nodes each, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) SBM-WS with 53 communities of size 151 nodes each, each community has the following WS parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithms are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The SBM-WS has similar clustering and modularity as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering or modularity that drives the formation of echo chambers. Other topologies must play a role as well. The most significant difference between the real-world Last.fm network and the SBM-WS lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution.\\ % add parameters of SBM-WS power??
\newline
It is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The Last.fm on-line social network is compared to network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The Last.fm on-line social network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The dotted black line represents a constant amount of echo chambers over time. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\newpage
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease. The increase is the strongest for the network models with high clustering (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. This seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering. \\
\newline
Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The SBM and SBM-WS only have communities of a single, fixed size. This does not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_REC.png}
    \label{pr_pow_single}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_PR.png}
    \label{rec_pow_single}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. The SBM-WS with heterogeneous community sizes (SBM-WS1) is compared to the SBM-WS with a single community size (SBM-WS2). Results are compared to the real-world Last.fm network with similar clustering and modularity. The dotted black line represents a constant amount of echo chambers over time. The distribution is normalized by dividing for the same quantity computed at $t=0$. Each curve is the average of $50$ independent simulations. Error bars are omitted for improved visualization.}
\label{power_vs_single}
\end{figure}

@@

Results for the PGP trust network are compared to results for network models with similar average degree, clustering or modularity. These models are, all with $N = 10680$: (i) WS model with $\left<k\right> = 4$, $\beta = 0.2$ and $\left<cc\right> = 0.266 \pm 0.003$; (ii) SBM with 60 communities of size 178 nodes each, $p_\text{cl} = 0.02$, $p_\text{add} = 0.00005$, $\left<k\right> = 4.06 \pm 0.02$ and $Q = 0.868 \pm 0.002$; (iii) SBM-WS with 60 communities of size 178 nodes each, each community has the following WS parameters: $\left<k\right> = 4$, $\beta = 0.12$; edges are added between communities with probability $p_\text{add} = 0.00007$, $\left<k\right> = 4.74 \pm 0.01$, $\left<cc\right> = 0.263 \pm 0.003$ and $Q = 0.842 \pm 0.002$.\\
\newline
In Fig. \ref{echo_vs_fracRes_real_PGP} the formation of echo chambers versus the fraction of stubborn actors is studied for the real-world PGP trust network and for the network models. This is done for both the PR and REC filtering algorithms. \\
\newline
The PR filtering algorithm (Fig. \ref{pr_real_PGP}) shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors for both the PGP trust network and the network models. The real-world PGP network has a significantly smaller formation of echo chambers compared to the network model with similar clustering and modularity (SBM-WS). This is most likely due to the heterogeneous degree pattern of the PGP network.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_PGP_8x7.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_PGP_8x7.png}
    \label{rec_real_PGP}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The PGP trust network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The PGP trust network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The dotted black line represents a constant amount of echo chambers over time. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real_PGP}
\end{figure}
\noindent
The REC filtering algorithm (Fig. \ref{rec_real_PGP}) shows an increase in the formation of echo chambers with an increasing fraction of stubborn actors, followed by a decrease if the fraction of stubborn actors is further increased. This is visible for both the PGP trust network and the network models. The PGP network has a smaller formation of echo chambers than the network model with similar clustering and modularity (SBM-WS), due to its heterogeneous degree distribution. The data points for the SBM and the PGP trust network coincide, which may suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering.

@@

We made use of simple toy models of social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of several layers, including a network layer and an opinion dynamics layer. Simple filtering algorithms were implemented to replicate and simulate the content curation used by social media companies. Furthermore, stubbornness was introduced to study the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B). We studied two starting conditions: balanced and unbalanced starting conditions.\\

@@

When introducing stubbornness, we found that the interplay between the PR filtering algorithm and individual resistance resulted in a decrease in the formation of echo chambers for the both the high clustering (WS, SBM-WS) and the high modularity networks (SBM, SBM-WS). The REF and REC filtering algorithms showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if the stubbornness was further increased. This is mainly seen in networks with high clustering, however a small bump can also be observed in networks with high modularity. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF and REC filtering algorithms and stubbornness is seen for both balanced and unbalanced starting conditions. It was found that including stubborn actors hampers the increase in the majority opinion (and thus prevents the formation of a final consensus state) in case of unbalanced starting conditions combined with the PR filtering algorithm.\\

@@

These results give us insight in the interplay between network structure, filtering algorithms and individual resistance. These insights may be crucial to understand phenomena such as polarization in society and to determine the impact of social media on these phenomena. They learn us that even though semantic filtering (PR) enhances the formation of echo chambers in the absence of stubborn actors, this enhancement does not stand on its own. An interplay between networks with high clustering or community structure is needed to effectively see the increase in echo chambers induced by semantic filtering. Random networks, combined with semantic filtering, do not enhance the formation of echo chambers. If there are no stubborn actors, recent or random filtering (REC or REF) seem to be more protective against the formation of echo chambers than semantic filtering. If there is a small fraction of stubborn actors, this scenario changes: semantic filtering starts to hamper the formation of echo chambers, whereas recent or random filtering begins to enhance polarization. It is often assumed that algorithmic personalization, tailored to ones preferences (e.g. semantic filtering), is one of the reasons for the appearance of echo chambers on social media \cite{Ge2020}\cite{Mohseni2018}\cite{Stark2020}. Our results may, however, suggest, that if there is stubbornness (which is probably the case in real life), heterogeneous exposure (REF or REC) to different opinions may be more harmful than homogeneous exposure (PR). Bail et al. found experimental results that may agree with this finding \cite{Bail2018}. This is an interesting result, that should be studied more thoroughly.\\

>>>>

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, in order to investigate the interplay between filtering algorithms and network structure.

@@

\begin{figure}[H]
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_stubbornness_SBM_frac_compl_stubb_PR_10x8.png}
	\captionsetup{format=plain}
	\caption[Formation of echo chambers versus fraction of stubborn actors for the PR filtering algorithm and balanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line represents a constant amount of echo chambers over time. The PR filtering algorithm is used. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$.}
\label{echo_vs_frac_complStubb}
\end{figure}
\noindent
The ER model shows no appearance of echo chambers. Introducing a fraction of stubborn actors does not change this. The SBM with low modularity behaves similar as the ER model. The data points for the SBM with low modularity lie behind the ER data points and are not visible in Fig. \ref{echo_vs_frac_complStubb}. The other networks (WS, SBM-WS and SBM with high modularity) all show qualitatively the same behavior. If an increasing fraction of stubborn actors is introduced, the formation of echo chambers decreases.\\
\newline
The WS and SBM-WS networks have high clustering. For a small fraction of stubborn actors they show the largest echo chambers. The SBM with high modularity has smaller echo chambers than networks with high clustering, but still has some formation of echo chambers for $f_s \leqslant 0.1$. These results confirm that community structure enhances the formation of echo chambers compared to random networks, however not as significant as networks with high clustering.

@@

  \caption[Formation of echo chambers versus fraction of stubborn actors for the PR and REC filtering algorithms and unbalanced starting conditions.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for unbalanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. The $y$-axis shows the total number of like-minded neighborhoods (sum of the number of echo chambers around opinion A and the number of echo chambers around opinion B) at $t=500$ divided by the total number of like-minded neighborhoods at $t=0$. The dotted black line represents a constant amount of echo chambers over time. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$; these are small and not visible.}

@@

  \caption[Formation of echo chambers versus the resistance parameter of the actors, $r$, for unbalanced starting conditions. The PR and REC filtering algorithms are used.]{Formation of echo chambers versus the resistance parameter of the actors, $r$, for unbalanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. Results are shown for: (i) WS; (ii) SBM-WS; (iii) SBM and (iv) ER. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity configuration ($Q = 0.908 \pm 0.005$)) and a low modularity configuration ($Q = 0.220 \pm 0.007$). The dotted black line represents a constant amount of echo chambers over time. The $y$-axis shows the total number of like-minded neighborhoods (sum of the number of echo chambers around opinion A and the number of echo chambers around opinion B) at $t=500$ divided by the total number of like-minded neighborhoods at $t=0$. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$; these are small and not visible.}

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. $10 \%$ and $50 \%$ of the communities is given a single opinion, opinion A, while the remaining communities get an initial $50/50$ opinion distribution. The community distributed case is compared to the corresponding randomly distributed case ($55/45$ or $75/25$) for both a high modularity and a low modularity configuration. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. First row: $f_\text{A} = 0.1$: \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Second row: $f_\text{A} = 0.5$: \textbf{(c)} REC filtering algorithm; \textbf{(d)} PR filtering algorithm. The community distributed case is compared to the randomly distributed case for both a high and a low modularity configuration. The dotted black line represents a constant amount of echo chambers over time. The distribution is normalized by dividing for the same quantity computed at $t=0$. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}

@@

The formation of echo chambers versus the fraction of stubborn actors, $f_s$, in the Last.fm on-line social network is studied. Results are compared to results from network models, all with a size of $N = 8003$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM with 53 communities of size 151 nodes each, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) SBM-WS with 53 communities of size 151 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithms are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The SBM-WS has similar clustering and modularity as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering or modularity that drives the formation of echo chambers. Other topologies must play a role as well. The most significant difference between the real-world Last.fm network and the SBM-WS lies in the degree distribution: the Last.fm network has a heterogeneous instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to network models with similar clustering and modularity, but with a homogeneous degree distribution.\\ % add parameters of SBM-WS power??
\newline
It is interesting to notice that the Last.fm on-line social network coincides with the SBM with similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of clustering.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The Last.fm on-line social network is compared to network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The Last.fm on-line social network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\newpage
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease. The increase is the strongest for network models with high clustering (WS, SBM-WS). Even though the Last.fm network has similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. This seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering. \\
\newline
Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The SBM and SBM-WS only have communities of a single, fixed size. This does not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_REC.png}
    \label{pr_pow_single}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_PR.png}
    \label{rec_pow_single}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. The SBM-WS with heterogeneous community sizes (SBM-WS1) is compared to the SBM-WS with a single community size (SBM-WS2). Results are compared to the real-world Last.fm network with similar clustering and modularity. The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line represents a constant amount of echo chambers over time. Each curve is the average of $50$ independent simulations. Error bars are omitted for improved visualization.}
\label{power_vs_single}
\end{figure}

@@

Results for the PGP trust network are compared to results for network models with similar average degree, clustering or modularity. These models are, all with a size of $N = 10680$ nodes: (i) WS model with $\left<k\right> = 4$, $\beta = 0.2$ and $\left<cc\right> = 0.266 \pm 0.003$; (ii) SBM with 60 communities of size 178 nodes each, $p_\text{cl} = 0.02$, $p_\text{add} = 0.00005$, $\left<k\right> = 4.06 \pm 0.02$ and $Q = 0.868 \pm 0.002$; (iii) SBM-WS with 60 communities of size 178 nodes each, each community has WS parameters: $\left<k\right> = 4$, $\beta = 0.12$; edges are added between communities with probability $p_\text{add} = 0.00007$, $\left<k\right> = 4.74 \pm 0.01$, $\left<cc\right> = 0.263 \pm 0.003$ and $Q = 0.842 \pm 0.002$.\\
\newline
In Fig. \ref{echo_vs_fracRes_real_PGP} the formation of echo chambers versus the fraction of stubborn actors, $f_s$, is studied for the real-world PGP trust network and for the network models. This is done for both the PR and REC filtering algorithms. \\
\newline
The PR filtering algorithm (Fig. \ref{pr_real_PGP}) shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors for both the PGP trust network and the network models. The real-world PGP network has a significantly smaller formation of echo chambers compared to the network model with similar clustering and modularity (SBM-WS). This is most likely due to the heterogeneous degree pattern of the PGP network.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_PGP_8x7.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_PGP_8x7.png}
    \label{rec_real_PGP}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The PGP trust network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of stubborn actors, $f_s$, for balanced starting conditions. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The PGP trust network is compared to network models: (i) WS; (ii) SBM-WS and (iii) SBM. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B. The dotted black line represents a constant amount of echo chambers over time. Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real_PGP}
\end{figure}
\newpage
\noindent
The REC filtering algorithm (Fig. \ref{rec_real_PGP}) shows an increase in the formation of echo chambers with an increasing fraction of stubborn actors, followed by a decrease if the fraction of stubborn actors is further increased. This is visible for both the PGP trust network and the network models. The PGP network has a smaller formation of echo chambers than the network model with similar clustering and modularity (SBM-WS), due to its heterogeneous degree distribution. The data points for the SBM and the PGP trust network coincide, which may suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the hampering due to a lack of clustering.

@@

We made use of simple toy models of social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of several layers, including a network layer and an opinion dynamics layer. Filtering algorithms were implemented to replicate and simulate the content curation used by social media companies. Furthermore, stubbornness was introduced to study the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B). We studied two starting conditions: balanced and unbalanced starting conditions.\\

@@

When introducing stubbornness, we found that the interplay between the PR filtering algorithm and individual resistance resulted in a decrease in the formation of echo chambers for the both the high clustering (WS, SBM-WS) and the high modularity networks (SBM, SBM-WS). The REF and REC filtering algorithms showed an increase in the formation of echo chambers with increasing stubbornness, followed by a decrease if stubbornness was further increased. This is mainly seen in networks with high clustering, however a small bump can also be observed in networks with high modularity. This interplay between the PR filtering algorithm and individual resistance versus the interplay between the REF and REC filtering algorithms and stubbornness is seen for both balanced and unbalanced starting conditions. Furthermore, it was found that including stubborn actors hampers the increase in the majority opinion (and thus prevents the formation of a final consensus state) in case of unbalanced starting conditions combined with the PR filtering algorithm.\\

@@

These results give us insight in the interplay between network structure, filtering algorithms and individual resistance. These insights may be crucial to understand phenomena such as polarization in society and to determine the impact of social media on these phenomena. They learn us that even though semantic filtering (PR) enhances the formation of echo chambers in the absence of stubborn actors, this enhancement does not stand on its own. An interplay between networks with high clustering or community structure is needed to effectively see the increase in echo chambers induced by semantic filtering. Random networks, combined with semantic filtering, do not enhance the formation of echo chambers. If there are no stubborn actors, recent or random filtering (REC or REF) seem to be more protective against the formation of echo chambers than semantic filtering. If there is a small fraction of stubborn actors, this scenario changes: semantic filtering starts to hamper the formation of echo chambers, whereas recent or random filtering begins to enhance polarization. It is often assumed that algorithmic personalization, tailored to ones preferences (e.g. semantic filtering), is one of the reasons for the appearance of echo chambers on social media \cite{Ge2020}\cite{Mohseni2018}\cite{Stark2020}. Our results may, however, suggest, that if there is stubbornness (which is probably the case in real life), heterogeneous exposure (REF or REC) to different opinions may be more harmful than homogeneous exposure (PR). Bail et al. found experimental results that may agree with this finding \cite{Bail2018}. This is a result that should be studied more thoroughly.\\
